Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-06 ms.qmd
Clean up of discussion

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate]. Phenotypic plasticity can allow populations to persist in the face of climate extremes by matching phenotypes to fitness optima, although this is not always the case [@noble2019plastic; @radersma2020plasticity] and there may be limits and costs to plasticity that constrain its expression [@Dewitt1998; but see @Murren2015]. We show that plastic adjustments in physiological rates and changes in their relative variance differs across habitats. In freshwater habitats, acclimation responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiological rates, suggesting acclimation may not be as beneficial. Relative variance reductions are particularly pronounced in terrestrial ectotherms, and this coincides with a general inability to acclimate. Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold-acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely to provide adaptive benefits under climate change. Together, these findings highlight the vulnerability of terrestrial ectotherms to climate change. Theoretical models that capture the percentage changes in relative variances and acclimation responses we observe here are needed fully appreciate the biological consequences such changes have on ecological and evolutionary dynamics. It may be expected that reductions in variance of this magnitude impact the 'opportunity for selection' by reducing the strength of selection and/or impacting underlying genetic variance exposed to selection. Both are predicted to impact the capacity of populations to evolve under climate change [@hoffmann2011climate; @urban2023and].

### ***Plasticity and variance in physiological rates do not differ between life-stages***
Interestingly, there were no major differences in acclimation or reductions in physiological variance between life-stages across habitat types. Greater plasticity for early life-stages of development is expected to be important to increase resilience to the effects of climate change because early life-stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life]. However, our results highlight that early life stages are not likely to be protected from the impacts of climate change. Evidence across the literature is equivocal with respect to the extent to which one life-stage is more or less plastic than another [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*. Across a diversity of taxa, we show that there were generally similar patterns between early and late life stages -- both in terms of variance changes and the capacity for plasticity. Unsurprisingly, there is much variation in the literature and our findings suggest that patterns will depend on the species and traits in question. However, very few studies measured multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same populations. A better understanding of the microthermal environments experienced by different life-stages will also be important in understanding why plasticity may or may not differ among life-stages.

@@

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were key traits most strongly impacted by increased temperature. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity. However, data testing the impacts of changes in variance on populations are limited to very few taxa. We need creative experimentation across broader taxa that manipulate levels of variance to understand their ecological outcomes. Our meta-analytic findings can be integrated into theoretical models by permitting model parameters to capture biologically relevant variation. Such predictive models can then be integrated with experimental approaches to better understand the consequences of changes in variation on ecological dynamics.

>>>>

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate]. Phenotypic plasticity can allow populations to persist in the face of climate extremes by matching phenotypes to fitness optima, although this is not always the case [@noble2019plastic; @radersma2020plasticity] and there may be limits and costs to plasticity that constrain its expression [@Dewitt1998; but see @Murren2015]. We show that plastic adjustments in physiological rates and changes in their relative variance differ across habitats. In freshwater habitats, acclimation responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiological rates, suggesting acclimation may not be as beneficial. Relative variance reductions are particularly pronounced in terrestrial ectotherms, and this coincides with a general inability to acclimate. Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold-acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely to provide adaptive benefits under climate change. Together, these findings highlight the vulnerability of terrestrial ectotherms to climate change. Theoretical models that capture the percentage changes in relative variances and acclimation responses we observe here are needed fully appreciate the biological consequences such changes have on ecological and evolutionary dynamics. It may be expected that reductions in variance of this magnitude impact the 'opportunity for selection' by reducing the strength of selection and/or impacting underlying genetic variance exposed to selection. Both are predicted to impact the capacity of populations to evolve under climate change [@hoffmann2011climate; @urban2023and].

### ***Plasticity and variance in physiological rates do not differ between life-stages***
Interestingly, there were no differences in acclimation or reductions in physiological variance between life-stages across habitat types. Greater plasticity for early life-stages of development is expected to be important to increase resilience to the effects of climate change because early life-stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life]. However, our results highlight that early life stages are not likely to be protected from the impacts of climate change. Evidence across the literature is equivocal with respect to the extent to which one life-stage is more or less plastic than another [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*. Across a diversity of taxa, we show that there were generally similar patterns between early and late life stages -- both in terms of variance changes and the capacity for plasticity. Unsurprisingly, there is much variation in the literature and our findings suggest that patterns will depend on the species and traits in question. However, very few studies measure multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same population.

@@

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were key traits most strongly impacted by increased temperature. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity in terrestrial ecosystems. However, data testing the impacts of changes in variance on populations are limited to very few taxa. We need creative experimentation across broader taxa that manipulate levels of variance to understand their ecological outcomes. Our meta-analytic findings can be integrated into theoretical models by permitting model parameters to capture biologically relevant changes in variation. Such predictive models can then be integrated with experimental approaches to better understand the consequences of changes in variation on ecological dynamics.
